---
title: "Processing of 3D Lidar data to assess forest structure"
author: "Svenja Dobelmann"
---

# Introduction
In this tutorial we will learn how to work with LiDAR data to estimate forest structure. To do so we will compare two forest plots in Lower Saxony of varying structure: One in the Göttinger Forest and the other in the Solling. Though both plots are dominated by beech stands of similar age, they differ in structural properties like tree height, canopy cover and understory regeneration. By looking into the vertical distribution of lidar returns, we can capture these differences. 



# Before we start
For this tutorial we will need several R packages that might need to be installed first.  

For handling Lidar Data we use the "lidR" package. A very good documentation of the package including further functionalities can be found here: <https://r-lidar.github.io/lidRbook/>

### Install and load required libraries
```{r, message=FALSE,results='hide'}


# load libraries
library(terra)        # Geospatial data (raster)
library(sf)           # Geospatial data (vector)
library(lidR)         # Laserscanning Data (lidar) 
library(rTwig)        # Quantitative Structure Models 
library(ggplot2)      # Pretty plotting
library(patchwork)    # plot arrangement 

# Or install (if required)
# install.packages("terra")
# install.packages("sf")
# install.packages("lidR", repos = c("https://r-lidar.r-universe.dev", "https://cloud.r-project.org"))
# install.packages("rTwig")
# install.packages("ggplot2")
# install.packages("patchwork")

```

We will work with two small point clouds (60x60m) acquired by Drone. To start with, let's download and import the two datasets.   

### Download data 
```{r, results='hide', eval=FALSE}
# Data access

## Download Pointcloud Göttingen
url_las_goe <- "https://cloud.hawk.de/index.php/s/GgTdjGMMz4rFoYR/download"
download.file(url_las_goe, destfile = "uls_goewald.laz", mode = "wb")

## Download pointcloud Solling 
url_las_sol <- "https://cloud.hawk.de/index.php/s/S923yWMyg6mdNFF/download"
download.file(url_las_sol, destfile = "uls_solling.laz", mode = "wb")



```
### Import the .laz files
```{r, results = 'hide'}

# load lidar point cloud  
las_goe <- readLAS("uls_goewald.laz")
las_sol <- readLAS("uls_solling.laz")
```

# Inspecting the point cloud data 

Before we compute anything, let's see what we are actually working with. Run the chunks below and answer the following:

1. In which format is the dataset stored?  
2. What are the pulse density and the point density, and why do the two differ?  
3. Which point attributes are stored in the datasets?  
4. Where do the two files differ?  
  
```{r}
## Göttinger Wald
print(las_goe)
## Solling 
print(las_sol)

```
```{r}

names(las_goe)

table(las_goe$NumberOfReturns)

```
## The LAS format

`print()` reveals that the two files are not stored identically: the Göttingen
data are **LAS version 1.4, point data record format 7**, the Solling data
**LAS version 1.2, point data record format 3**. Both numbers matter:

- The **version** defines the structure of the file as a whole — its
  header, the coordinate scaling, and which record formats are permitted.
- The **point data record format** (PDRF) defines which attributes are stored for
  each individual point.

What exactly changes between versions and formats is laid out in the
specification, which is maintained by the **American Society for Photogrammetry
and Remote Sensing (ASPRS)** — also the reason the classification codes
(2 = ground, 5 = high vegetation, …) are known as ASPRS classes.

Full specification:
<https://publicdocuments.asprs.org/las-v14-r16-2025>


## Pointcloud visualisation
`plot()` opens an interactive 3D window in which the point cloud can be
rotated and zoomed. Colouring the points by height makes the vertical
structure easier to read.

```{r, eval=FALSE}
plot(las_goe, color = "Z", bg = "white")
plot(las_sol, color = "Z", bg = "white")
```

Based on this first visual inspection, what structural differences can you see between the two stands?  

### Cutting a transect
The 3D view is useful for a first impression, but it is hard to compare
two stands in a rotating window, so let's cut a slice through the pointcloud and project it onto a plane, so we can see the vertical profile. We'll use a little helper function that cuts a 5 m wide slice along the centre line of each point cloud. 

```{r}
# helper function for transect-clipping
get_transect <- function(las) {
  b <- st_bbox(las)
  cy <- (b$ymin + b$ymax) / 2
  clip_transect(las, c(b$xmin, cy), c(b$xmax, cy), width = 5, xz = TRUE)
}

# clipping the transects 
las_tr_goe <- get_transect(las_goe)
las_tr_sol <- get_transect(las_sol)

print(las_tr_goe)

```

### Plotting the transects

```{r, warning = FALSE, fig.width=10, fig.height=3.8}
## plot the transects with ggplot
p_goe <- ggplot(payload(las_tr_goe), aes(X,Z)) + 
  geom_point(size = 0.5, color = "black") + 
  labs(title = "Göttinger Forest",
       x = "X",
       y = "Height (ASL)"
       ) + 
  coord_equal() + 
  theme_minimal() 

p_sol <- ggplot(payload(las_tr_sol), aes(X,Z)) + 
  geom_point(size = 0.5, color = "black") + 
  labs(title = "Solling Forest",
       x = "X",
       y = "Height (ASL)"
  ) +
  coord_equal() + 
  theme_minimal() 

# side by side plotting using patchwork 
p_goe | p_sol
```
The profiles show the vertical structure, but the Z values are still heights above sea level (ASL). Together with the sloping terrain it is impossible to compare the actual tree heights between the two sites.   


Normalisation solves this: we subtract the terrain height from every point, so that Z becomes height above ground. `normalize_height()` overwrites the original Z value with the height above Ground (HAG)

The function needs two things: which classification code identifies the ground points (2 under the ASPRS scheme), and an algorithm to interpolate a continuous terrain surface from them. Here we use `tin()`, a Delaunay triangulation of the ground returns, from which the height beneath each point is then subtracted.

### Normalizing Height 
```{r, results = 'hide'}
# normalize height 
las_goe_norm <- normalize_height(las = las_goe, 
                        algorithm = tin(), 
                        use_class = 2)


las_sol_norm <- normalize_height(las = las_sol, 
                            algorithm = tin(), 
                            use_class = 2)

```

With the normalised heights in place, we can cut the transects again and plot HAG instead of Z. The ground now forms a flat line at zero, and the vertical axis reads directly as tree height.  


Compare the two profiles again. Which differences that were hidden by the terrain are now visible, and how tall are the tallest trees in each stand?

```{r, warning = FALSE, fig.width=10, fig.height=3.8}

las_tr_goe <- get_transect(las_goe_norm)
las_tr_sol <- get_transect(las_sol_norm)

p_goe <- ggplot(payload(las_tr_goe), aes(X,Z)) + 
  geom_point(size = 0.5, color = "black") + 
  labs(title = "Göttinger Forest",
       x = "X",
       y = "Height Above Ground (HAG)"
       ) + 
  ylim(c(0,40)) + 
  coord_equal() + 
  theme_minimal() 

p_sol <- ggplot(payload(las_tr_sol), aes(X,Z)) + 
  geom_point(size = 0.5, color = "black") + 
  labs(title = "Solling Forest",
       x = "X",
       y = "Height Above Ground (HAG)"
  ) +
  ylim(c(0,40)) + 
  coord_equal() + 
  theme_minimal() 

# side by side plotting 
p_goe | p_sol
```

### Add vertical distribution of points 
To quantify these structural differences, we can look into the vertical distribution of the returns. Let's add a density curve beside each transect that shows how the points are spread over height. We keep only points above ground, since ground returns would otherwise produce a spike at zero that dominates the curve.


```{r, warning = FALSE}
# Height distributions
h_goe <- ggplot(subset(payload(las_goe_norm), Z > 0), aes(Z)) +
  geom_density(fill = "grey70", color = NA) +
  xlim(0, 40) +
  coord_flip() +
  theme_void()

h_sol <- ggplot(subset(payload(las_sol_norm), Z > 0), aes(Z)) +
  geom_density(fill = "grey70", color = NA) +
  xlim(0, 40) +
  coord_flip() +
  theme_void()

# Combine
(p_goe | h_goe) + plot_layout(widths = c(5, 1)) 
(p_sol | h_sol) + plot_layout(widths = c(5, 1))
```

## Calculating Canopy Height Models 

So far we have looked at the point cloud from the side. Now let's look at it
from above. A **Canopy Height Model (CHM)** is a raster in which every cell
holds the height of the highest point within it.   

This only works because we normalised the point clouds earlier. Since Z is now
height above ground, the raster values can be read directly as tree height.  

**`res = 0.5`** sets the cell size in metres. The resolution should match the
point density  

**`dsmtin()`** builds the surface by Delaunay triangulation of the first
returns, the same principle we used for the terrain model in
`normalize_height()`, only applied to the top of the canopy instead of the
ground. Each raster cell then takes its value from the triangulated surface.  



```{r, fig.width=10, fig.height=5, warning = FALSE}

chm_goe <- rasterize_canopy(las_goe_norm, res = 0.5, algorithm = dsmtin())

chm_sol <- rasterize_canopy(las_sol_norm, res = 0.5, algorithm = dsmtin())

par(mfrow = c(1,2))
plot(chm_goe, main = "CHM Göttingen")
plot(chm_sol, main = "CHM Solling")
```

Compare the two rasters and consider:  

1. Where is the canopy closed, and where do you see gaps reaching down towards
   the ground?  
2. How variable is the canopy surface within each stand? A rough, uneven surface
   points to trees of differing age and size, a smooth one to a uniform stand.  
3. Can you identify individual crowns? In which of the two stands is this easier,
   and why?  

### Metrics at the plot level 
We can also put numbers to this visual impression. `cloud_metrics()` calculates
a set of summary statistics for the whole point cloud. `.stdmetrics_z` is lidR's
standard collection of height metrics. Two filters are applied before calculating the metrics: Only **first returns** are used and points below zero are dropped. 


```{r}
# first returns only, and drop points below the interpolated ground surface
prep <- function(las){
  filter_poi(filter_first(las), Z >= 0)
             }

# calculate metrics
m_goe <- cloud_metrics(prep(las_goe_norm), .stdmetrics_z)
m_sol <- cloud_metrics(prep(las_sol_norm), .stdmetrics_z)

# here we use a selection of metrics only 
vars <- c("zmax", "zmean", "zsd", "zq95", "pzabove2", "pzabovezmean", "zentropy")

# combine results 
round(rbind(Göttinger = unlist(m_goe)[vars],
            Solling   = unlist(m_sol)[vars]), 2)

```
Are these results what you expected after looking at the transects? Which stand
appeared structurally more diverse there, and which one comes out on top here?



## Individual Tree Detection

So far we have described the stands as a whole. The next step is to resolve
them into individual trees. 

This happens in two steps. **Individual Tree Detection (ITD)** locates the
position and height of each treetop. **Individual Tree Segmentation (ITS)** then
assigns every return in the point cloud to one of those trees.

### The local maximum filter 
The standard approach is a **Local Maximum Filter (LMF)**. A window is moved
across the canopy surface, and wherever the centre cell is the highest point
within that window, a tree top is recorded. The window size is a crucial parameter here.  

Since crown size scales with tree height, we let the window grow with the height of the point being tested. we calculate window size as a function of height. 

```{r, fig.width=5, fig.height=4}
# variable window size as a function of height
f <- function(x) {
  y <- 2.6 * (-(exp(-0.08 * (x - 2)) - 1)) + 3
  # after https://r-lidar.github.io/lidRbook/itd.html
  y[x < 2]  <- 3
  y[x > 20] <- 5
  return(y)
}

heights <- seq(0, 40, 0.5)

plot(heights, f(heights), type = "l", ylim = c(0, 6),
     xlab = "Height above ground (m)", ylab = "Window size (m)",
     main = "Variable window size")
```
### Running the tree detection

```{r, results = 'hide'}
ttops_goe <- locate_trees(chm_goe, lmf(ws = f, hmin = 5))
ttops_sol <- locate_trees(chm_sol, lmf(ws = f, hmin = 5))

```
`hmin = 5` sets a detection floor at 5 m, so that understorey vegetation and
noise near the ground are not counted as trees. The result is an `sf` object
with one point per tree, carrying its coordinates and its height. We can add the tree tops on top of the Canopy Height model to visually inspect the results.   

Which plot has more trees in total? 
```{r, fig.width=10, fig.height=5, warning = F}
par(mfrow = c(1, 2))
plot(chm_goe, main = "Göttinger Forest")
plot(st_geometry(ttops_goe), add = TRUE, pch = 3, cex = 0.6)
plot(chm_sol, main = "Solling")
plot(st_geometry(ttops_sol), add = TRUE, pch = 3, cex = 0.6)
```

## Individual Tree Segmentation

Detection gives us points. Segmentation gives us trees: every return in the
cloud is assigned to one of the detected treetops.

`dalponte2016()` works on the raster. Starting from each treetop, it grows a
region outwards into neighbouring cells as long as they stay within a plausible
height range for that crown, and stops where the surface drops off or another
crown is reached. `segment_trees()` then transfers the resulting crown outlines
back to the point cloud and writes a new attribute, `treeID`, into every point.  

```{r}
# segment point cloud
las_goe_norm <- segment_trees(las_goe_norm, dalponte2016(chm_goe, ttops_goe)) 
las_sol_norm <- segment_trees(las_sol_norm, dalponte2016(chm_sol, ttops_sol)) 

```

Colouring the point cloud by `treeID` shows how well the delineation worked.
```{r, eval=FALSE}
# 3D plot
x <- plot(las_goe_norm, bg = "white", size = 4,color = "treeID")
add_treetops3d(x, ttops_goe)

x <- plot(las_sol_norm, bg = "white", size = 4,color = "treeID")
add_treetops3d(x, ttops_sol)
```
Did the tree segmentation work equally well in both plots? If not, what could be the reason for the differences? 

### Metrics at the tree level
Because every point now knows which tree it belongs to, we can derive attributes
per tree rather than per plot:

```{r}
crowns_goe <- crown_metrics(las_goe_norm, func = .stdtreemetrics)
crowns_sol <- crown_metrics(las_sol_norm, func = .stdtreemetrics)

crowns_goe
crowns_sol

```
This gives us the height of each tree and the area of each tree crown. 

## Forest structural complexity: the box dimension

Every metric so far captured one aspect of structure, and each ranked our two
stands differently. The **box dimension (Db)** condenses all of it into a single
number.

The point cloud is divided into cubic voxels, and the occupied ones are counted.
The voxels are then made progressively smaller and the count repeated. Db is the
slope of the regression between log box count and log inverse box size: 
A structure that is simple and open gains little from finer inspection and
yields a low value. One that fills space with detail at every scale — many
crowns, several layers, branches occupying the gaps between them — keeps
producing newly occupied voxels and yields a high one. The R² of the fit
indicates how consistently that behaviour repeats across scales.


The concept is covered in detail in the accompanying document on fractal
complexity — here we simply apply it.  


first we filter only points above 0,5 m height, to exclude ground and ground vegetation.  
```{r}
# X, Y, Z of all points above 0.5 m, from the normalised clouds
cloud_goe <- las_goe_norm@data[Z > 0.5, 1:3]
cloud_sol <- las_sol_norm@data[Z > 0.5, 1:3]
```

`lowercutoff = 0.1` sets the smallest voxel edge to 10 cm — below that the count
would describe the scanner's sampling pattern rather than the forest.
`rm_int_box = TRUE` drops the single box enclosing the whole cloud, which
carries no structural information but would leverage the slope.

```{r, fig.width=5, fig.height=5}

db_goe <- box_dimension(cloud_goe, lowercutoff = 0.1, rm_int_box = TRUE, plot = "2D")
db_sol <- box_dimension(cloud_sol, lowercutoff = 0.1, rm_int_box = TRUE, plot = "2D")

```




Which stand has the higher box dimension, and does the ranking agree with
`zentropy` and the transect profiles?

One caveat: the method was developed for terrestrial scans at centimetre
resolution. Our drone data are denser than most ALS but far from that standard,
and the closed Göttinger canopy occludes its understorey more than the Solling
does. The comparison between our two plots holds — same sensor, same
configuration — but the absolute values should not be compared to published
TLS-based figures.  